\documentclass[11pt]{article}

\input{../../../_shared/notes-common.tex}

\title{Elements of AIML Lab\\Experiment 06: Missing Values \& Outliers (Student Handout — Before Submission)}
\author{Tofik Ali}
\date{\today}

\begin{document}
\maketitle
\tableofcontents

\section{About this handout}
This is the \textbf{pre-submission} student handout.
It intentionally provides only \textbf{requirements} and \textbf{hints}.
After submission, you may receive a separate \textbf{reference} document for cross-checking.

\section{Objective}
\begin{itemize}[leftmargin=*]
  \item Detect missing values and outliers in a dataset.
  \item Apply and compare multiple handling strategies and document the impact.
\end{itemize}

\section{What you must do (minimum requirements)}
\begin{enumerate}[leftmargin=*]
  \item Use the dataset from Experiment~04/05 (preferred) or another approved dataset.
  \item Report missing values per column (count and \%).
  \item Apply at least \textbf{two} missing-value strategies (e.g., drop vs impute) and compare outcomes.
  \item Detect outliers for at least \textbf{one} numeric column using a rule such as \textbf{IQR} or \textbf{z-score}.
  \item Handle outliers using at least \textbf{one} approach (remove or cap/winsorize) and justify.
  \item Compare before vs after using summaries and at least one visual (recommended: boxplot).
  \item Write 3--5 observations describing what changed and why your choices are reasonable.
\end{enumerate}

\section{Hints (read only if stuck)}
\begin{itemize}[leftmargin=*]
  \item Typical Pandas helpers: \texttt{df.isna().sum}, \texttt{df.isna().mean}, \texttt{df.describe}, \texttt{fillna}, \texttt{dropna}.
  \item Median imputation is often safer than mean on skewed data.
  \item Outliers can be errors or rare-but-real events; do not delete blindly.
  \item If you later train ML models, avoid leakage: fit imputers/rules on training data only.
\end{itemize}

\section{What to submit (MANDATORY)}
Submit a single zipped folder containing:
\begin{itemize}[leftmargin=*]
  \item \texttt{notebook.ipynb}: steps performed, before/after comparisons, and observations.
  \item \texttt{README.md}: dataset link/source, strategies used, and justification summary.
  \item \texttt{assets/} (optional): plot images if saved separately.
\end{itemize}
Dataset rule: do not upload large datasets; provide a link/source instead.

\section{Submission checklist}
\begin{itemize}[leftmargin=*]
  \item Missingness reported (count and \%) for columns.
  \item Two missing-value strategies shown and compared.
  \item One outlier rule applied + one handling approach applied + justification written.
  \item Before/after impact shown with summaries and at least one plot.
  \item Notebook + README included and readable.
\end{itemize}

\section{Common mistakes}
\begin{itemize}[leftmargin=*]
  \item Deleting many rows without explaining the cost/impact.
  \item Handling missing values/outliers but not showing before/after evidence.
  \item Applying rules on the full dataset in an ML setting (leakage risk).
\end{itemize}

\notesources{\AIMLLabSources}
\end{document}
