\documentclass[aspectratio=169]{beamer}

\input{../../../_shared/beamer-common.tex}

\title[Elements of AIML]{Elements of AIML}
\subtitle{Unit 03 -- Lecture 03: Dimensionality Reduction (PCA, LDA, ICA)}
\author{Tofik Ali}
\institute{School of Computer Science, UPES Dehradun}
\date{\today}

\graphicspath{{../images/}{../../../_shared/}{../../../_shared/UPES-Beamer-Theme/}}

\begin{document}

\UPESCoverFrame
\setcounter{framenumber}{0}

\begin{frame}
  \titlepage
  \vspace{0.3em}
  \begin{center}
    \footnotesize Topic focus: \textbf{reduce dimensions} for speed, noise reduction, and visualization\\[0.25em]
    \footnotesize Repository: \texttt{\AIMLRepoURL}
  \end{center}
  \slidesources{\AIMLUnitIIISources}
\end{frame}

\begin{frame}{Quick Links}
  \centering
  \hyperlink{sec:why}{\beamerbutton{Why Reduce?}}\hspace{0.6em}
  \hyperlink{sec:pca}{\beamerbutton{PCA}}\hspace{0.6em}
  \hyperlink{sec:ldaica}{\beamerbutton{LDA/ICA}}\hspace{0.6em}
  \hyperlink{sec:example}{\beamerbutton{Example}}\hspace{0.6em}
  \hyperlink{sec:summary}{\beamerbutton{Summary}}
\end{frame}

\begin{frame}{Agenda}
  \tableofcontents
\end{frame}

\section{Why reduce dimensions?}
\label{sec:why}

\begin{frame}{Learning Outcomes}
  By the end of this lecture, you should be able to:
  \begin{itemize}[<+->]
    \item Explain why high-dimensional data can be difficult (curse of dimensionality).
    \item Describe PCA intuition (maximum variance directions).
    \item Outline PCA steps and interpret explained variance.
    \item Distinguish PCA vs LDA vs ICA at a conceptual level.
  \end{itemize}
\end{frame}

\begin{frame}{Abbreviations \& Symbols}
  \begin{itemize}[<+->]
    \item $n$: samples,\; $d$: original features,\; $k$: reduced dimensions
    \item $X \in \mathbb{R}^{n\times d}$: data matrix,\; $Z \in \mathbb{R}^{n\times k}$: reduced data
    \item $C$: covariance matrix,\; $\lambda_i$: eigenvalues,\; $v_i$: eigenvectors
    \item PC: principal component,\; EVR: explained variance ratio
  \end{itemize}
\end{frame}

\begin{frame}{Curse of Dimensionality (Intuition)}
  \begin{itemize}[<+->]
    \item In high dimensions, data becomes sparse.
    \item Distances can become less informative.
    \item Models may overfit and require more data.
  \end{itemize}
  \vspace{0.4em}
  Dimensionality reduction helps with \textbf{compression}, \textbf{noise reduction}, and \textbf{visualization}.
\end{frame}

\section{PCA}
\label{sec:pca}

\begin{frame}{PCA: Intuition}
  \begin{block}{Principal Component Analysis (PCA)}
    Find orthogonal directions (components) that capture \textbf{maximum variance} in the data.
  \end{block}
  \vspace{0.4em}
  \begin{itemize}[<+->]
    \item First component: direction with highest variance.
    \item Next components: highest remaining variance, orthogonal to previous.
  \end{itemize}
\end{frame}

\begin{frame}{PCA Geometry (2D Intuition)}
  \centering
  \begin{tikzpicture}[scale=0.85]
    \draw[->] (-0.2,0) -- (5.2,0) node[right] {$x_1$};
    \draw[->] (0,-0.2) -- (0,4.2) node[above] {$x_2$};
    \draw[thick, blue!70] (0.6,0.6) -- (4.4,3.8) node[right] {PC1};
    \foreach \x/\y in {0.8/0.9,1.2/1.1,1.7/1.5,2.2/1.9,2.6/2.2,3.0/2.6,3.4/3.0,3.8/3.2,4.1/3.5}{
      \fill[black] (\x,\y) circle (1.4pt);
    }
  \end{tikzpicture}
  \vspace{0.3em}
  \small PCA chooses the direction with maximum spread (variance) and projects data onto it.
\end{frame}

\begin{frame}{PCA Steps (High Level)}
  \begin{enumerate}[<+->]
    \item Standardize features (often needed).
    \item Compute covariance matrix.
    \item Compute eigenvectors/eigenvalues.
    \item Choose top-$k$ eigenvectors (components).
    \item Project data onto these components.
  \end{enumerate}
\end{frame}

\begin{frame}{Key Formulas (What PCA computes)}
  \begin{itemize}[<+->]
    \item Centered data: $X_c = X - \mu$
    \item Covariance: $C = \frac{1}{n-1} X_c^\top X_c$
    \item Eigen-decomposition: $C v_i = \lambda_i v_i$
    \item Projection (keep top-$k$): $Z = X_c V_k$
  \end{itemize}
  \vspace{0.3em}
  \small $V_k$ contains the top-$k$ eigenvectors (principal directions).
\end{frame}

\begin{frame}{Explained Variance}
  \begin{itemize}[<+->]
    \item Eigenvalues indicate how much variance each component captures.
    \item Explained variance ratio helps choose $k$.
    \item Goal: keep enough components to retain most information.
  \end{itemize}
  \vspace{0.4em}
  Example rule: choose the smallest $k$ with $\ge 90\%$ explained variance (depends on use-case).
\end{frame}

\begin{frame}{Choosing $k$ (Scree Plot Idea)}
  \begin{itemize}[<+->]
    \item Plot EVR per component: large drop $\Rightarrow$ ``elbow'' point.
    \item Or use cumulative EVR: pick smallest $k$ reaching a target (e.g., 90\%).
    \item Trade-off: compression vs information loss.
  \end{itemize}
\end{frame}

\section{LDA and ICA (overview)}
\label{sec:ldaica}

\begin{frame}{PCA vs LDA vs ICA (Concept)}
  \begin{itemize}[<+->]
    \item \textbf{PCA}: unsupervised; maximizes variance.
    \item \textbf{LDA}: supervised; finds projections that separate classes.
    \item \textbf{ICA}: unsupervised; finds statistically independent components (source separation idea).
  \end{itemize}
\end{frame}

\begin{frame}{Quick Comparison Table}
  \centering
  \begin{tabular}{llll}
    \toprule
    \textbf{Method} & \textbf{Uses labels?} & \textbf{Goal} & \textbf{Typical use} \\
    \midrule
    PCA & No & maximize variance & compression/visualization \\
    LDA & Yes & separate classes & classification features \\
    ICA & No & independent sources & signal separation \\
    \bottomrule
  \end{tabular}
\end{frame}

\section{Mini example}
\label{sec:example}

\begin{frame}{Mini Example (2D to 1D Projection)}
  \begin{itemize}[<+->]
    \item Suppose points lie roughly along a diagonal line in 2D.
    \item PCA finds the diagonal direction and projects data to 1D.
    \item Result: compressed representation with minimal information loss.
  \end{itemize}
  \vspace{0.4em}
  In the lab, you will implement PCA and visualize explained variance.
\end{frame}

\begin{frame}[fragile]{Minimal PCA in Python (sketch)}
  \begin{lstlisting}[language=Python]
from sklearn.preprocessing import StandardScaler
from sklearn.decomposition import PCA

X_scaled = StandardScaler().fit_transform(X)
pca = PCA(n_components=2)
Z = pca.fit_transform(X_scaled)
print(pca.explained_variance_ratio_)
  \end{lstlisting}
  \small Standardize first; then inspect explained variance ratios.
\end{frame}

\section{Summary}
\label{sec:summary}

\begin{frame}{Summary}
  \begin{itemize}
    \item Dimensionality reduction combats high-dimensional issues and improves efficiency.
    \item PCA finds maximum-variance orthogonal directions; explained variance guides $k$.
    \item LDA uses labels to separate classes; ICA seeks independent components.
  \end{itemize}
  \slidesources{\AIMLUnitIIISources}
\end{frame}

\begin{frame}{Exit Questions}
  \begin{enumerate}
    \item What does PCA optimize?
    \item Why do we often standardize before PCA?
    \item Compare PCA and LDA in one paragraph.
    \item What does ``explained variance'' mean?
  \end{enumerate}
\end{frame}

\end{document}
