\documentclass[11pt]{article}

\input{../../../_shared/notes-common.tex}

\title{Elements of AIML Lab\\Experiment 05: EDA in Python (Student Handout — Before Submission)}
\author{Tofik Ali}
\date{\today}

\begin{document}
\maketitle
\tableofcontents

\section{About this handout}
This is the \textbf{pre-submission} student handout.
It intentionally provides only \textbf{requirements} and \textbf{hints}.
After submission, you may receive a separate \textbf{reference} document for cross-checking.

\section{Objective}
\begin{itemize}[leftmargin=*]
  \item Perform exploratory data analysis (EDA) to understand data quality and patterns.
  \item Create a minimum set of plots and write observations based on evidence.
  \item Recommend preprocessing actions for later experiments.
\end{itemize}

\section{What you must do (minimum requirements)}
\begin{enumerate}[leftmargin=*]
  \item Use the dataset from Experiment~04 (preferred) or another instructor-approved dataset.
  \item Report basic structure: rows/columns, types, missing values, duplicates (if any).
  \item Produce at least \textbf{three} plots:
  \begin{itemize}[leftmargin=*]
    \item one distribution plot (e.g., histogram/KDE) for a numeric feature,
    \item one boxplot for a numeric feature,
    \item one relationship plot (e.g., scatter plot or correlation heatmap).
  \end{itemize}
  \item For each plot, write 2--3 lines of observation (what you see and what it suggests).
  \item Write an EDA summary with 5--8 bullet observations and 2--3 recommended preprocessing actions.
\end{enumerate}

\section{Hints (read only if stuck)}
\begin{itemize}[leftmargin=*]
  \item Typical Pandas functions used in EDA: \texttt{df.info}, \texttt{df.describe}, \texttt{df.isna().sum}, \texttt{df.duplicated().sum}.
  \item Typical plotting tools: Matplotlib + Seaborn (choose plot type based on variable types).
  \item EDA is incomplete without written observations (plots alone are not enough).
\end{itemize}

\section{What to submit (MANDATORY)}
Submit a single zipped folder containing:
\begin{itemize}[leftmargin=*]
  \item \texttt{notebook.ipynb}: EDA notebook with plots and observations.
  \item \texttt{README.md}: dataset link/source, list of plots produced, and a short EDA summary.
  \item \texttt{assets/} (optional): exported plot images if you saved them separately.
\end{itemize}
Dataset rule: do not upload large datasets; provide a link/source instead.

\section{Submission checklist}
\begin{itemize}[leftmargin=*]
  \item At least 3 required plot types included with observations.
  \item Missing values and duplicates are reported.
  \item EDA summary includes actions recommended for later labs.
  \item Notebook + README included and readable.
\end{itemize}

\section{Common mistakes}
\begin{itemize}[leftmargin=*]
  \item Creating plots without any written interpretation.
  \item Using wrong plot types (e.g., histogram for categories).
  \item Not checking missingness/duplicates before plotting and summarizing.
\end{itemize}

\notesources{\AIMLLabSources}
\end{document}
