\documentclass[11pt]{article}

\input{../../../_shared/notes-common.tex}

\title{Elements of AIML Lab\\Experiment 05 (After-submission Reference): EDA using Python}
\author{Tofik Ali}
\date{\today}

\begin{document}
\maketitle
\tableofcontents

\section{How to use this reference}
This reference is meant to be shared \textbf{after you submit} Experiment~05.
It demonstrates a complete EDA flow and, more importantly, shows \textbf{how to write observations} (EDA is not only plots).

\section{EDA in one sentence}
EDA answers: \emph{``What does this dataset look like, what can go wrong, and what should we do next?''}

\section{Recommended EDA sequence (teaching order)}
\subsection{1) Load + understand structure}
\begin{lstlisting}[language=Python]
import pandas as pd

df = pd.read_csv("processed.csv")   # use Exp-04 output if available
print(df.shape)
print(df.head())
print(df.info())
\end{lstlisting}
\textbf{What to write:} mention rows/columns and the main column types (numeric/categorical/date).

\subsection{2) Data quality checks (missing + duplicates)}
\begin{lstlisting}[language=Python]
miss = df.isna().sum().sort_values(ascending=False)
print(miss[miss > 0])
print("Duplicates:", df.duplicated().sum())
\end{lstlisting}
\textbf{What to write:} name the top 2--3 columns with missing values and their counts/percentages.

\subsection{3) Descriptive summaries}
\begin{lstlisting}[language=Python]
print(df.describe())
print(df.describe(include="object"))
\end{lstlisting}
\textbf{What to look for:} min/max surprises, skewed distributions, rare categories.

\subsection{4) Univariate plots (distribution of one feature)}
\begin{lstlisting}[language=Python]
import matplotlib.pyplot as plt
import seaborn as sns

sns.histplot(df["num_col"], kde=True)
plt.title("Distribution of num_col")
plt.show()

sns.boxplot(x=df["num_col"])
plt.title("Boxplot of num_col")
plt.show()
\end{lstlisting}
\textbf{Observation examples:}
\begin{itemize}[leftmargin=*]
  \item ``\texttt{num\_col} is right-skewed; most values are below 50 with a long tail up to 300.''
  \item ``The boxplot shows extreme values; we may need outlier handling or log transform.''
\end{itemize}

\subsection{5) Relationship plots (feature vs feature)}
\begin{lstlisting}[language=Python]
corr = df.select_dtypes("number").corr()
sns.heatmap(corr, cmap="coolwarm", center=0)
plt.title("Correlation heatmap (numeric features)")
plt.show()
\end{lstlisting}
\textbf{How to interpret:}
\begin{itemize}[leftmargin=*]
  \item strong positive correlation $\approx$ features increase together,
  \item strong negative correlation $\approx$ one increases as the other decreases,
  \item near zero correlation does not guarantee independence (could be non-linear).
\end{itemize}

\subsection{6) If there is a target column (optional but valuable)}
\begin{itemize}[leftmargin=*]
  \item Numeric target: scatter plot feature vs target.
  \item Class label: compare distributions by class (boxplots or histograms per class).
\end{itemize}

\section{Minimum plot set (good standard)}
\begin{itemize}[leftmargin=*]
  \item One histogram (distribution) + written comment.
  \item One boxplot (outliers/spread) + written comment.
  \item One relationship plot (scatter or heatmap) + written comment.
\end{itemize}

\section{Common EDA writing pattern (use this in reports)}
For each plot/check, write:
\begin{enumerate}[leftmargin=*]
  \item \textbf{Evidence:} what you saw (numbers/plot pattern).
  \item \textbf{Meaning:} what it suggests about the data.
  \item \textbf{Action:} what you would do next (missing handling, scaling, encoding, etc.).
\end{enumerate}

\section{Troubleshooting}
\begin{itemize}[leftmargin=*]
  \item \textbf{Plots look empty:} check column names, data types, and missing values.
  \item \textbf{Too many categories:} plot top-$k$ value counts and group rare categories as ``Other''.
  \item \textbf{Time columns:} convert with \texttt{pd.to\_datetime} before time-based plots.
\end{itemize}

\notesources{\AIMLLabSources}
\end{document}

