\documentclass[11pt]{article}

\input{../../../_shared/notes-common.tex}

\title{Elements of AIML Lab\\Experiment 02: WEKA Classification (Student Handout — Before Submission)}
\author{Tofik Ali}
\date{\today}

\begin{document}
\maketitle
\tableofcontents

\section{About this handout}
This is the \textbf{pre-submission} student handout.
It intentionally provides only \textbf{requirements} and \textbf{hints}.
After submission, you may receive a separate \textbf{reference} document for cross-checking.

\section{Objective}
\begin{itemize}[leftmargin=*]
  \item Perform a classification task in WEKA (predict a \textbf{label} / \textbf{nominal} class).
  \item Train and compare at least \textbf{two} classifiers using a correct evaluation method.
  \item Report the confusion matrix and key metrics (precision/recall/F1) and write observations.
\end{itemize}

\section{What you must do (minimum requirements)}
\begin{enumerate}[leftmargin=*]
  \item Choose a dataset where the class attribute is \textbf{nominal} (categorical label).
  \item Load the dataset in \textbf{WEKA Explorer} and set the \textbf{Class} attribute correctly.
  \item Train at least \textbf{two} classification models in the \textbf{Classify} tab.
  \item Evaluate using \textbf{10-fold cross-validation} \emph{or} a \textbf{percentage split} (same setting for both models).
  \item Record results for each model:
  \begin{itemize}[leftmargin=*]
    \item confusion matrix,
    \item precision/recall/F1 (per-class or weighted average, as applicable).
  \end{itemize}
  \item Write 3--5 observations (trade-offs and which model you prefer, with reasons).
\end{enumerate}

\section{Hints (read only if stuck)}
\begin{itemize}[leftmargin=*]
  \item If the class attribute is numeric, you are not doing classification.
  \item Avoid \textbf{Use training set} for evaluation.
  \item Use the same test option for all models (fair comparison).
  \item If classes are imbalanced, do not rely only on accuracy; focus on precision/recall/F1 and the confusion matrix.
  \item Common classifiers you may try: \texttt{J48}, \texttt{NaiveBayes}, \texttt{Logistic}, \texttt{SMO}, \texttt{RandomForest}.
\end{itemize}

\section{What to submit (MANDATORY)}
Submit a single zipped folder containing:
\begin{itemize}[leftmargin=*]
  \item \texttt{notebook.ipynb}: your lab record (can be markdown-focused) with screenshots and a comparison table.
  \item \texttt{README.md}: dataset link/source, class labels, models compared, evaluation method, and observations.
  \item \texttt{assets/} (folder): screenshots/images used in the notebook/README.
\end{itemize}
Dataset rule: do not upload large datasets; provide a link/source instead.

\section{Submission checklist}
\begin{itemize}[leftmargin=*]
  \item Class attribute is correct and nominal.
  \item Two classifiers compared using the same evaluation method.
  \item Confusion matrix + precision/recall/F1 included.
  \item Notebook + README included and readable.
\end{itemize}

\section{Common mistakes}
\begin{itemize}[leftmargin=*]
  \item Wrong class attribute (predicting an ID or unrelated column).
  \item Comparing models using different test options.
  \item Reporting accuracy only (especially when data is imbalanced).
  \item Missing evidence of WEKA settings/outputs.
\end{itemize}

\notesources{\AIMLLabSources}
\end{document}
