\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{listings}

\geometry{margin=1in}
\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  urlcolor=blue
}

\setlist{itemsep=0.35em, topsep=0.35em}

\pagestyle{fancy}
\fancyhf{}
\lhead{Elements of AIML (B.Tech)}
\rhead{Unit 01: Introduction to AI}
\cfoot{\thepage}
\setlength{\headheight}{14pt}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  stringstyle=\color{teal},
  showstringspaces=false,
  columns=fullflexible,
  frame=single
}

\newcommand{\SubmissionDeadline}{March 8, 2026}

\begin{document}

\begin{center}
  {\LARGE \textbf{Assignment -- Unit 01}}\\[0.25em]
  {\large Elements of AIML}\\[0.25em]
  \normalsize (AI overview, applications, techniques, success criteria, intelligent agents, environments, learning agents, limitations)
\end{center}

\noindent
\textbf{Instructor:} Tofik Ali \hfill \textbf{Submission Deadline:} \SubmissionDeadline\\
\textbf{Student Name:} \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_ \hfill \textbf{Roll No.:} \_\_\_\_\_\_\_\_\\

\vspace{0.6em}
\hrule
\vspace{0.8em}

\textbf{Instructions}
\begin{itemize}[leftmargin=*]
  \item Answer \textbf{all} questions. Use your own examples and clearly state assumptions.
  \item Submission format: submit a single ZIP containing:
    \begin{itemize}[leftmargin=*]
      \item \texttt{notebook.ipynb} (typed answers in Markdown + equations; insert diagrams/screenshots if used)
      \item \texttt{README.md} (1-page summary: what you learned + references/links you used)
      \item optional \texttt{assets/} folder for images
    \end{itemize}
  \item Write concise answers, but do not skip reasoning. Bullet points are allowed.
  \item Use simple, clear language (assume a non-expert reader).
\end{itemize}

\vspace{0.6em}

\section*{Easy (10 Questions)}
\begin{enumerate}[label=E\arabic*., leftmargin=*]
  \item \textbf{Define AI:} Give a 2--3 line definition of Artificial Intelligence in your own words.
    Mention \textbf{one} key goal (e.g., decision-making, perception, language, planning).

  \item \textbf{AI applications:} List \textbf{five} AI applications across different domains.
    For each, write: \emph{input} $\rightarrow$ \emph{output} (one line each).

  \item \textbf{AI vs ML vs Data Science:} Write 6--10 lines comparing \textbf{AI}, \textbf{Machine Learning}, and \textbf{Data Science}.
    Give one example project for each.

  \item \textbf{Techniques (high-level):} Name \textbf{three} broad categories of AI techniques and give one example of each:
    \begin{enumerate}[label=(\alph*), leftmargin=*]
      \item Search / planning
      \item Knowledge / logic (reasoning)
      \item Learning (data-driven)
    \end{enumerate}

  \item \textbf{Criteria of success:} For each scenario, write \textbf{two} success criteria beyond accuracy:
    \begin{enumerate}[label=(\alph*), leftmargin=*]
      \item A medical triage assistant
      \item A spam filter
      \item A campus navigation app
      \item An AI-based loan approval support tool
    \end{enumerate}

  \item \textbf{Agent basics:} What is an \textbf{agent}? Give one example and list its \textbf{sensors} and \textbf{actuators}.

  \item \textbf{PEAS:} Create a PEAS description for a \textbf{spam filter}.
    (Performance measure, Environment, Actuators, Sensors)

  \item \textbf{Environment properties (intro):} Classify each environment as best as you can:
    \emph{(fully/partially observable), (deterministic/stochastic), (episodic/sequential), (static/dynamic)}.
    \begin{enumerate}[label=(\alph*), leftmargin=*]
      \item Chess
      \item Online shopping recommendation
    \end{enumerate}
    Justify in 2--3 lines each.

  \item \textbf{Rationality:} Define a \textbf{rational agent} in 3--5 lines.
    Explain the difference between a \textbf{performance measure} and a \textbf{reward} (one paragraph).

  \item \textbf{Limitations:} List \textbf{four} real-world limitations of AI systems (deployment-time issues).
    For each, write one possible mitigation idea (one line).
\end{enumerate}

\section*{Medium (10 Questions)}
\begin{enumerate}[label=M\arabic*., leftmargin=*]
  \item \textbf{PEAS (design):} Create a PEAS description for an \textbf{online exam proctoring system}.
    Include at least 5 items for Performance measure and Sensors.

  \item \textbf{Environment classification:} Consider a \textbf{smart thermostat} in a hostel room.
    Classify the environment using at least \textbf{five} properties (observable, deterministic, episodic, static, discrete, etc.).
    Explain in 6--10 lines.

  \item \textbf{Agent types:} Compare (in a table) the following agent types:
    \textbf{simple reflex}, \textbf{model-based reflex}, \textbf{goal-based}, \textbf{utility-based}, \textbf{learning agent}.
    For each, write one example task.

  \item \textbf{Choosing an agent:} You must build an \textbf{elevator controller} for a 10-floor academic building.
    Choose one agent type from M3 and justify your choice in 8--12 lines.

  \item \textbf{System success vs model accuracy:} A face-recognition attendance system reports 95\% accuracy,
    but students complain about frequent false rejections for some groups.
    Write 8--12 lines explaining why \emph{system success} can still be low and what you would measure instead.

  \item \textbf{Bias sources:} List three ways bias can enter an AI system (data, measurement, deployment, etc.).
    For each, write a practical mitigation.

  \item \textbf{Sensors matter:} Give one example where adding a new sensor changes the environment from \emph{less observable} to \emph{more observable}.
    Explain how it changes the agent design (6--10 lines).

  \item \textbf{Applications + risks table:} Create a table with 6 AI domains (e.g., healthcare, finance, education, security, retail, transport).
    For each: one typical task + one major risk/failure mode.

  \item \textbf{Limitations in deployment:} Write a short ``deployment checklist'' (8--12 bullets) for releasing an AI feature in a university app.
    Include: data privacy, monitoring, human override, and failure handling.

  \item \textbf{Impact:} Choose one domain (education / healthcare / banking / security).
    Write 10--14 lines covering: benefits, harms, stakeholders, and what ``responsible use'' means.
\end{enumerate}

\section*{Hard (10 Questions)}
\begin{enumerate}[label=H\arabic*., leftmargin=*]
  \item \textbf{Full case design:} Design an AI system for \textbf{waste collection optimization} in a city.
    Provide: agent description, PEAS, success criteria, and 3 key limitations/risks with mitigations (1--2 pages).

  \item \textbf{Self-driving environment analysis:} Explain why self-driving is a difficult AI task using environment properties.
    Classify it using at least 6 properties and justify each (12--18 lines). Mention sensors/actuators briefly.

  \item \textbf{Learning agent decomposition:} For a \textbf{news recommendation system}, describe the components of a learning agent:
    performance element, learning element, critic, problem generator.
    Draw a simple block diagram and explain each component's role (12--18 lines).

  \item \textbf{Failure postmortem:} An AI chatbot for admissions gave incorrect scholarship rules.
    Write a postmortem report (1 page) including: what failed, why, user impact, and a prevention plan (data + process + monitoring).

  \item \textbf{Rule-based vs learning-based:} Compare rule-based and learning-based approaches for a spam filter.
    Give one situation where a rule-based approach is still preferable and justify (10--14 lines).

  \item \textbf{Evaluation plan:} Propose an evaluation plan for an AI feature that suggests course electives to students.
    Include offline evaluation, user feedback, monitoring after launch, and a rollback plan (10--16 bullets).

  \item \textbf{Ethics and governance:} Create a 1-page ``Responsible AI'' policy note for a student-facing AI tool.
    Include fairness, privacy, transparency, accountability, and safety.

  \item \textbf{Explainability scenario:} Choose one AI application where explainability is required.
    Explain (10--14 lines) what kind of explanation is needed and for whom (student/admin/regulator).

  \item \textbf{AI limitations case:} Give a real or realistic example of an AI system failing due to distribution shift (data changes over time).
    Explain what changed and how to detect/handle it (12--18 lines).

  \item \textbf{Mini reflection:} Write a short reflection (12--18 lines):
    ``Where is AI overhyped today, and where is it genuinely useful?'' Support with examples.
\end{enumerate}

\vfill
\noindent\textit{End of Assignment}

\end{document}
