\documentclass[11pt]{article}

\input{../../../_shared/notes-common.tex}

\title{Elements of AIML Lab\\Experiment 08: PCA (Student Handout — Before Submission)}
\author{Tofik Ali}
\date{\today}

\begin{document}
\maketitle
\tableofcontents

\section{About this handout}
This is the \textbf{pre-submission} student handout.
It intentionally provides only \textbf{requirements} and \textbf{hints}.
After submission, you may receive a separate \textbf{reference} document for cross-checking.

\section{Objective}
\begin{itemize}[leftmargin=*]
  \item Reduce dimensionality using PCA and justify the number of components.
  \item Visualize explained variance and (if applicable) a 2D PCA projection.
\end{itemize}

\section{What you must do (minimum requirements)}
\begin{enumerate}[leftmargin=*]
  \item Use a dataset with multiple numeric features (from previous labs or approved dataset).
  \item Apply PCA to numeric features (do not include labels/targets as PCA input features).
  \item Report explained variance ratio and cumulative explained variance.
  \item Choose a number of components $k$ and justify it (e.g., 90\% variance or elbow point).
  \item Produce:
  \begin{itemize}[leftmargin=*]
    \item one cumulative explained variance plot,
    \item one 2D scatter plot of PC1 vs PC2 (if possible for your dataset).
  \end{itemize}
  \item Write 3--5 observations (compression achieved, trade-offs, and when PCA is useful).
\end{enumerate}

\section{Hints (read only if stuck)}
\begin{itemize}[leftmargin=*]
  \item Standardize numeric features before PCA if they are on different scales (\texttt{StandardScaler}).
  \item PCA tools: \texttt{PCA}, \texttt{explained\_variance\_ratio\_}, cumulative sum.
  \item If you later train a model, avoid leakage: fit scaler+PCA on training data only (pipeline recommended).
\end{itemize}

\section{What to submit (MANDATORY)}
Submit a single zipped folder containing:
\begin{itemize}[leftmargin=*]
  \item \texttt{notebook.ipynb}: PCA steps, plots, chosen $k$, and observations.
  \item \texttt{README.md}: dataset link/source, chosen $k$ with justification, and key results.
  \item \texttt{assets/} (optional): plot images if saved separately.
\end{itemize}
Dataset rule: do not upload large datasets; provide a link/source instead.

\section{Submission checklist}
\begin{itemize}[leftmargin=*]
  \item PCA applied on numeric features (no label leakage into PCA inputs).
  \item EVR + cumulative EVR reported and plotted.
  \item $k$ chosen and justified.
  \item Notebook + README included and readable.
\end{itemize}

\section{Common mistakes}
\begin{itemize}[leftmargin=*]
  \item Skipping standardization when features have very different scales.
  \item Choosing $k$ without justification or without an EVR plot.
  \item Including the target/label column in PCA inputs.
\end{itemize}

\notesources{\AIMLLabSources}
\end{document}
