\documentclass[11pt]{article}

\input{../../../_shared/notes-common.tex}

\title{Elements of AIML Lab\\Experiment 08 (After-submission Reference): PCA (Dimensionality Reduction)}
\author{Tofik Ali}
\date{\today}

\begin{document}
\maketitle
\tableofcontents

\section{How to use this reference}
This reference is meant to be shared \textbf{after you submit} Experiment~08.
It clarifies:
\begin{itemize}[leftmargin=*]
  \item why scaling is needed before PCA,
  \item how to choose number of components using explained variance,
  \item how to avoid leakage when PCA is used before a model.
\end{itemize}

\section{PCA: what it learns}
PCA finds new axes (principal components) such that:
\begin{itemize}[leftmargin=*]
  \item PC1 captures maximum variance,
  \item PC2 captures maximum remaining variance subject to being orthogonal to PC1,
  \item and so on.
\end{itemize}
Each principal component is a weighted combination of original features.

\section{Worked example (end-to-end)}
\subsection{Step 1: load data and select features}
\begin{lstlisting}[language=Python]
import pandas as pd

df = pd.read_csv("processed.csv")  # or your dataset
X = df.select_dtypes("number")     # PCA requires numeric features
\end{lstlisting}
If you have a label column, keep it separately (do not include in PCA features).

\subsection{Step 2: scale + fit PCA}
\begin{lstlisting}[language=Python]
import numpy as np
from sklearn.preprocessing import StandardScaler
from sklearn.decomposition import PCA

scaler = StandardScaler()
X_scaled = scaler.fit_transform(X)

pca = PCA()
Z = pca.fit_transform(X_scaled)

evr = pca.explained_variance_ratio_
cum = np.cumsum(evr)
print("First 10 EVR:", evr[:10])
print("First 10 cumulative:", cum[:10])
\end{lstlisting}

\subsection{Step 3: pick number of components}
Two common choices:
\begin{itemize}[leftmargin=*]
  \item choose the smallest $k$ such that cumulative EVR $\ge 0.90$ (90\% variance),
  \item or choose $k$ at the ``elbow'' of the cumulative curve (diminishing returns).
\end{itemize}
You can also let scikit-learn choose by ratio:
\begin{lstlisting}[language=Python]
pca90 = PCA(n_components=0.90)
Z90 = pca90.fit_transform(X_scaled)
print("Chosen components:", pca90.n_components_)
\end{lstlisting}

\subsection{Step 4: plot cumulative explained variance}
\begin{lstlisting}[language=Python]
import matplotlib.pyplot as plt

plt.plot(cum, marker="o")
plt.axhline(0.90, color="red", linestyle="--", label="90%")
plt.xlabel("Number of components")
plt.ylabel("Cumulative explained variance")
plt.grid(True)
plt.legend()
plt.show()
\end{lstlisting}

\subsection{Step 5: interpret components (loadings)}
Loadings show which original features contribute to each PC.
\begin{lstlisting}[language=Python]
import pandas as pd

loadings = pd.DataFrame(
    pca.components_.T,
    index=X.columns,
    columns=[f"PC{i+1}" for i in range(pca.n_components_)]
)

print(loadings["PC1"].sort_values(key=abs, ascending=False).head(8))
\end{lstlisting}
\textbf{Interpretation tip:} large magnitude loading $\Rightarrow$ that feature strongly influences that component.

\section{Leakage-safe PCA (important)}
If you use PCA before training a model:
\begin{itemize}[leftmargin=*]
  \item split train/test first,
  \item fit scaler+PCA on training only,
  \item transform test using fitted objects,
  \item ideally use a pipeline.
\end{itemize}
Example:
\begin{lstlisting}[language=Python]
from sklearn.pipeline import Pipeline
from sklearn.linear_model import LogisticRegression

pipe = Pipeline([
    ("scaler", StandardScaler()),
    ("pca", PCA(n_components=0.90)),
    ("clf", LogisticRegression(max_iter=1000)),
])
\end{lstlisting}

\section{Troubleshooting}
\begin{itemize}[leftmargin=*]
  \item \textbf{PCA fails:} check for non-numeric columns and missing values.
  \item \textbf{EVR is low:} your features may be weakly correlated; PCA may not compress much.
  \item \textbf{PC plot looks noisy:} PCA is not a classifier; separation is not guaranteed.
\end{itemize}

\notesources{\AIMLLabSources}
\end{document}

