\documentclass[11pt]{article}

\input{../../../_shared/notes-common.tex}

\title{Elements of AIML Lab\\Experiment 09: Imbalanced Data Handling (Student Handout — Before Submission)}
\author{Tofik Ali}
\date{\today}

\begin{document}
\maketitle
\tableofcontents

\section{About this handout}
This is the \textbf{pre-submission} student handout.
It intentionally provides only \textbf{requirements} and \textbf{hints}.
After submission, you may receive a separate \textbf{reference} document for cross-checking.

\section{Objective}
\begin{itemize}[leftmargin=*]
  \item Show why accuracy can be misleading on imbalanced datasets.
  \item Apply at least one imbalance-handling technique and compare results.
\end{itemize}

\section{What you must do (minimum requirements)}
\begin{enumerate}[leftmargin=*]
  \item Use an imbalanced classification dataset (minority class must be clearly smaller).
  \item Report class distribution (counts and \%).
  \item Train a baseline classifier and report:
  \begin{itemize}[leftmargin=*]
    \item confusion matrix,
    \item precision/recall/F1 (and accuracy if you want, but not alone).
  \end{itemize}
  \item Apply at least \textbf{one} technique (choose any one):
  \begin{itemize}[leftmargin=*]
    \item class weights,
    \item oversampling/undersampling (train only),
    \item SMOTE (optional; if library available),
    \item threshold tuning (optional).
  \end{itemize}
  \item Compare baseline vs improved approach using the same split and metrics.
  \item Write 3--5 observations explaining what improved and what trade-off you see.
\end{enumerate}

\section{Hints (read only if stuck)}
\begin{itemize}[leftmargin=*]
  \item Always use stratified train/test split for imbalanced data.
  \item Resampling must be done on \textbf{training data only} (otherwise it leaks test information).
  \item A common simple approach is \texttt{class\_weight="balanced"} (many models support it).
  \item Expect trade-offs: improving recall can reduce precision (more false alarms).
\end{itemize}

\section{What to submit (MANDATORY)}
Submit a single zipped folder containing:
\begin{itemize}[leftmargin=*]
  \item \texttt{notebook.ipynb}: baseline vs improved comparison with metrics and observations.
  \item \texttt{README.md}: dataset link/source, technique chosen, and summary of results/trade-offs.
  \item \texttt{assets/} (optional): any saved outputs/plots.
\end{itemize}
Dataset rule: do not upload large datasets; provide a link/source instead.

\section{Submission checklist}
\begin{itemize}[leftmargin=*]
  \item Class distribution shown (counts and \%).
  \item Baseline metrics included (confusion matrix + precision/recall/F1).
  \item One imbalance-handling technique applied and compared fairly.
  \item Notebook + README included and readable.
\end{itemize}

\section{Common mistakes}
\begin{itemize}[leftmargin=*]
  \item Reporting accuracy only.
  \item Resampling before the split (leakage).
  \item Changing multiple things at once and not being able to explain the effect.
\end{itemize}

\notesources{\AIMLLabSources}
\end{document}
