\documentclass[11pt]{article}

\input{../../../_shared/notes-common.tex}

\title{Elements of AIML Lab\\Experiment 09 (After-submission Reference): Imbalanced Classification}
\author{Tofik Ali}
\date{\today}

\begin{document}
\maketitle
\tableofcontents

\section{How to use this reference}
This reference is meant to be shared \textbf{after you submit} Experiment~09.
It explains why accuracy fails on imbalanced data and shows standard improvement methods with correct evaluation practice.

\section{The core problem}
If positives are rare (e.g., 1\% fraud), a model can get 99\% accuracy by predicting ``not fraud'' always.
Therefore, for rare positives we usually focus on:
\begin{itemize}[leftmargin=*]
  \item recall (catching positives),
  \item precision (avoiding false alarms),
  \item F1, PR-AUC (optional but strong),
  \item confusion matrix (always).
\end{itemize}

\section{Worked example (self-contained)}
If you used a real dataset, your numbers differ. This example creates an imbalanced dataset to demonstrate the workflow.

\subsection{Step 1: create imbalanced data}
\begin{lstlisting}[language=Python]
import numpy as np
from sklearn.datasets import make_classification
from sklearn.model_selection import train_test_split

X, y = make_classification(
    n_samples=4000,
    n_features=10,
    n_informative=4,
    n_redundant=2,
    weights=[0.97, 0.03],   # 3% positive class
    random_state=42,
)

X_train, X_test, y_train, y_test = train_test_split(
    X, y, test_size=0.2, random_state=42, stratify=y
)
\end{lstlisting}

\subsection{Step 2: baseline model + evaluation}
\begin{lstlisting}[language=Python]
from sklearn.linear_model import LogisticRegression
from sklearn.metrics import confusion_matrix, classification_report

base = LogisticRegression(max_iter=2000)
base.fit(X_train, y_train)
pred = base.predict(X_test)

print(confusion_matrix(y_test, pred))
print(classification_report(y_test, pred, digits=3))
\end{lstlisting}
Interpret the confusion matrix: if FN is high, the model is missing positives.

\subsection{Step 3A: class weights (simple and effective)}
\begin{lstlisting}[language=Python]
weighted = LogisticRegression(max_iter=2000, class_weight="balanced")
weighted.fit(X_train, y_train)
pred_w = weighted.predict(X_test)

print(confusion_matrix(y_test, pred_w))
print(classification_report(y_test, pred_w, digits=3))
\end{lstlisting}
Typical outcome: recall increases, precision may decrease (more false alarms). Explain the trade-off.

\subsection{Step 3B: resampling (train only)}
If you have \texttt{imbalanced-learn} installed:
\begin{lstlisting}[language=Python]
from imblearn.over_sampling import SMOTE

sm = SMOTE(random_state=42)
X_res, y_res = sm.fit_resample(X_train, y_train)
print("Before:", np.bincount(y_train), "After:", np.bincount(y_res))
\end{lstlisting}
Then train on \texttt{X\_res, y\_res} and evaluate on the original test set.
\textbf{Rule:} never resample before the train/test split.

\subsection{Step 4 (optional): probability scores + PR-AUC}
\begin{lstlisting}[language=Python]
from sklearn.metrics import average_precision_score

scores = weighted.predict_proba(X_test)[:, 1]
ap = average_precision_score(y_test, scores)
print("PR-AUC (Average Precision):", round(ap, 4))
\end{lstlisting}

\section{What strong observations look like}
\begin{itemize}[leftmargin=*]
  \item ``Baseline accuracy was high, but recall for the minority class was near zero, so it missed most positives.''
  \item ``Using class\_weight increased recall from 0.10 to 0.72 with a moderate precision drop; this is acceptable when FN is costly.''
  \item ``Resampling improved recall further but increased false positives; threshold tuning may be needed depending on domain costs.''
\end{itemize}

\section{Troubleshooting}
\begin{itemize}[leftmargin=*]
  \item \textbf{No \texttt{predict\_proba}:} some models require probability-enabled settings.
  \item \textbf{Scores look perfect:} check for leakage (resampling before split, or target leakage features).
  \item \textbf{Unstable results:} use stratified splits and fix \texttt{random\_state}.
\end{itemize}

\notesources{\AIMLLabSources}
\end{document}

