\documentclass[11pt]{article}

\input{../../../_shared/notes-common.tex}

\title{Elements of AIML Lab\\Experiment 03: WEKA Clustering (Student Handout — Before Submission)}
\author{Tofik Ali}
\date{\today}

\begin{document}
\maketitle
\tableofcontents

\section{About this handout}
This is the \textbf{pre-submission} student handout.
It intentionally provides only \textbf{requirements} and \textbf{hints}.
After submission, you may receive a separate \textbf{reference} document for cross-checking.

\section{Objective}
\begin{itemize}[leftmargin=*]
  \item Perform clustering in WEKA (unsupervised learning).
  \item Run and compare at least \textbf{two} clusterers on the same dataset.
  \item Show the effect of feature scaling on distance-based clustering and write interpretations.
\end{itemize}

\section{What you must do (minimum requirements)}
\begin{enumerate}[leftmargin=*]
  \item Choose a dataset suitable for clustering (numeric features preferred).
  \item Load it in \textbf{WEKA Explorer} and remove obvious ID-like columns (if any).
  \item Run \textbf{k-means} (\texttt{SimpleKMeans}) and record cluster sizes and key outputs.
  \item Run a \textbf{second} clusterer (e.g., \texttt{EM} or \texttt{HierarchicalClusterer}) and record outputs.
  \item Apply a scaling/normalization filter and repeat clustering to compare results.
  \item Add 3--5 observations interpreting the clusters (what they might represent) and what changed after scaling.
\end{enumerate}

\section{Hints (read only if stuck)}
\begin{itemize}[leftmargin=*]
  \item Useful filters: \texttt{Normalize} or \texttt{Standardize}.
  \item If clusters look meaningless, re-check: scaling, removal of IDs, and the chosen number of clusters ($k$).
  \item Clusters are not ``true labels''; your job is to describe patterns and possible meanings.
  \item Include at least one visualization screenshot (cluster assignments) if available.
\end{itemize}

\section{What to submit (MANDATORY)}
Submit a single zipped folder containing:
\begin{itemize}[leftmargin=*]
  \item \texttt{notebook.ipynb}: your lab record (can be markdown-focused) with screenshots and a comparison table.
  \item \texttt{README.md}: dataset link/source, clusterers used, scaling used, and observations/interpretation.
  \item \texttt{assets/} (folder): screenshots/images used in the notebook/README.
\end{itemize}
Dataset rule: do not upload large datasets; provide a link/source instead.

\section{Submission checklist}
\begin{itemize}[leftmargin=*]
  \item Two clusterers run on the same dataset.
  \item One comparison included: before vs after scaling/normalization.
  \item Evidence included: settings + outputs + at least one visualization screenshot.
  \item Notebook + README included and readable.
\end{itemize}

\section{Common mistakes}
\begin{itemize}[leftmargin=*]
  \item Forgetting scaling (distance-based clustering can be distorted by feature units).
  \item Treating clusters as ground truth labels instead of exploratory patterns.
  \item Choosing $k$ without any justification or comparison.
  \item Missing screenshots of settings/outputs (cannot verify work).
\end{itemize}

\notesources{\AIMLLabSources}
\end{document}
