\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{listings}

\geometry{margin=1in}
\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  urlcolor=blue
}

\setlist{itemsep=0.35em, topsep=0.35em}

\pagestyle{fancy}
\fancyhf{}
\lhead{Elements of AIML (B.Tech)}
\rhead{Units 01--05: Diagnostic Assessment}
\cfoot{\thepage}
\setlength{\headheight}{14pt}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  stringstyle=\color{teal},
  showstringspaces=false,
  columns=fullflexible,
  frame=single
}

\newcommand{\SubmissionDeadline}{April 28, 2026}
\newcommand{\EMarks}{\textbf{[2 marks]} }
\newcommand{\MMarks}{\textbf{[3 marks]} }
\newcommand{\HMarks}{\textbf{[5 marks]} }

\begin{document}

\begin{center}
  {\LARGE \textbf{Diagnostic Assessment -- Units 01--05}}\\[0.25em]
  {\large Elements of AIML}\\[0.25em]
  \normalsize (Cumulative assessment for Unit 01 to Unit 05)
\end{center}

\noindent
\textbf{Instructor:} Tofik Ali \hfill \textbf{Time Limit:} None\\
\textbf{Submission Deadline:} \SubmissionDeadline\\
\textbf{Student Name:} \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_ \hfill \textbf{SAP-ID:} \_\_\_\_\_\_\_\_\_\_\_\_\\

\vspace{0.6em}
\hrule
\vspace{0.8em}

\textbf{Instructions}
\begin{itemize}[leftmargin=*]
  \item \textbf{This is a diagnostic assessment, not a speed test.} The goal is to identify conceptual gaps across \textbf{Unit 01 to Unit 05}.
  \item \textbf{Total marks: 200.} Easy: 20 $\times$ 2 = 40, Medium: 20 $\times$ 3 = 60, Hard: 20 $\times$ 5 = 100.
  \item \textbf{No time limit.} Submit on or before \textbf{\SubmissionDeadline}.
  \item \textbf{Important (70\% rule):} to have the linked assignment counted, the target score is at least 70\%,
    that is \textbf{140/200 or higher}. Reattempts are allowed.
  \item \textbf{Marking policy (honest attempt):} marks are deducted mainly for non-attempts and high similarity with other submissions.
    Correctness matters, but reasoning and visible effort matter more.
  \item Show key steps for logic and ML questions. If you are uncertain, write your assumptions and attempt anyway.
  \item Multiple valid answers are acceptable for open-ended design questions when they are logically consistent and technically justified.
  \item Unless stated otherwise, round numeric answers to \textbf{2 decimal places}.
\end{itemize}

\vspace{0.6em}

\section*{Easy (20 Questions)}
\begin{enumerate}[label=\textbf{E\arabic*.}, leftmargin=*]
  \item \EMarks \textbf{[Subjective]} Define AI in one sentence and mention one common deployment limitation.

  \item \EMarks \textbf{[MCQ -- Select all that apply]} Which are parts of PEAS?
    \begin{enumerate}[label=(\Alph*), leftmargin=*]
      \item Performance measure
      \item Environment
      \item Actuators
      \item Sensors
      \item Training dataset
    \end{enumerate}

  \item \EMarks \textbf{[Subjective]} Classify chess as fully or partially observable,
    and as deterministic or stochastic. Justify briefly.

  \item \EMarks \textbf{[MCQ -- Select one]} A rational agent chooses actions that:
    \begin{enumerate}[label=(\Alph*), leftmargin=*]
      \item maximize expected performance measure based on available percepts
      \item always maximize immediate reward only
      \item never use internal state
      \item solve every problem optimally in all environments
    \end{enumerate}

  \item \EMarks \textbf{[Subjective]} Name the four components of a learning agent and state the role of the critic.

  \item \EMarks \textbf{[MCQ -- Select one]} A sentence is \textbf{valid} if it is:
    \begin{enumerate}[label=(\Alph*), leftmargin=*]
      \item true in at least one model
      \item true in all models
      \item false in all models
      \item true only in the intended model
    \end{enumerate}

  \item \EMarks \textbf{[Subjective]} Write the truth table for $P \rightarrow Q$ and state when it is false.

  \item \EMarks \textbf{[Subjective]} Convert $P \rightarrow Q$ into an equivalent form using only $\lnot$ and $\lor$.

  \item \EMarks \textbf{[Subjective]} Explain $KB \models \alpha$ in plain words in 2--4 lines.

  \item \EMarks \textbf{[Subjective]} Perform one resolution step on:
    \[
      (P \lor Q),\ (\lnot P \lor R).
    \]

  \item \EMarks \textbf{[MCQ -- Select all that apply]} Which are quantifiers in first-order logic?
    \begin{enumerate}[label=(\Alph*), leftmargin=*]
      \item $\forall$
      \item $\exists$
      \item $\land$
      \item $\rightarrow$
      \item $\lor$
    \end{enumerate}

  \item \EMarks \textbf{[Subjective]} State the difference between a function symbol and a predicate symbol.

  \item \EMarks \textbf{[Subjective]} Define dataset, feature, and label using one small ML example.

  \item \EMarks \textbf{[MCQ -- Select one]} Which split should normally be touched only once at the end?
    \begin{enumerate}[label=(\Alph*), leftmargin=*]
      \item training set
      \item validation set
      \item test set
      \item bootstrap sample
    \end{enumerate}

  \item \EMarks \textbf{[Subjective]} Define data leakage and give one example.

  \item \EMarks \textbf{[Subjective]} State what PCA tries to optimize.

  \item \EMarks \textbf{[MCQ -- Select one]} Which metric is most natural for a continuous prediction problem?
    \begin{enumerate}[label=(\Alph*), leftmargin=*]
      \item recall
      \item F1-score
      \item MAE
      \item silhouette score
    \end{enumerate}

  \item \EMarks \textbf{[Subjective]} Why can accuracy be misleading for imbalanced classification?

  \item \EMarks \textbf{[MCQ -- Select one]} Clustering is mainly:
    \begin{enumerate}[label=(\Alph*), leftmargin=*]
      \item supervised learning
      \item unsupervised learning
      \item reinforcement learning
      \item online search
    \end{enumerate}

  \item \EMarks \textbf{[Subjective]} Give one responsible AI risk in an education application and one mitigation.
\end{enumerate}

\section*{Medium (20 Questions)}
\begin{enumerate}[label=\textbf{M\arabic*.}, leftmargin=*]
  \item \MMarks \textbf{[Subjective]} Give a PEAS description for a campus navigation assistant.

  \item \MMarks \textbf{[Subjective]} Give three success criteria beyond accuracy for an AI system in healthcare.

  \item \MMarks \textbf{[Subjective]} Let $KB=\{P \rightarrow Q,\ Q \rightarrow R,\ P\}$.
    Check whether $KB \models R$ by truth table/model checking.

  \item \MMarks \textbf{[Subjective]} Determine satisfiability of:
    \[
      (P \lor Q),\ (\lnot P \lor Q),\ (P \lor \lnot Q).
    \]
    Provide one satisfying assignment if it exists.

  \item \MMarks \textbf{[Subjective]} Convert
    \[
      (P \rightarrow Q) \land (Q \rightarrow R)
    \]
    to CNF.

  \item \MMarks \textbf{[Subjective]} Translate to propositional logic:
    ``If the server is down then send an alert. The server is down. Therefore send an alert.''

  \item \MMarks \textbf{[Subjective]} Translate into FOL:
    ``Every student studies at least one course.''

  \item \MMarks \textbf{[Subjective]} Find an MGU if it exists:
    \[
      Knows(John,x) \quad \text{and} \quad Knows(y,Bill).
    \]

  \item \MMarks \textbf{[Subjective]} Explain occurs-check using the attempt to unify $x$ and $f(x)$.

  \item \MMarks \textbf{[Subjective]} For a dataset of 100 rows, propose a train/validation/test split and explain your choice.

  \item \MMarks \textbf{[Subjective]} When do we use k-fold cross-validation and why is it useful?

  \item \MMarks \textbf{[Subjective]} PCA returns explained variance ratios
    \[
      [0.60, 0.20, 0.10, 0.06, 0.04].
    \]
    How many components are required to retain at least 85\% variance?

  \item \MMarks \textbf{[Subjective]} Compare PCA and LDA in 5--8 lines.

  \item \MMarks \textbf{[Subjective]} Distinguish regression and classification, and give one metric for each.

  \item \MMarks \textbf{[Subjective]} A classifier gives TP = 30, FP = 10, FN = 20, TN = 40.
    Compute precision and recall.

  \item \MMarks \textbf{[Subjective]} Why is accuracy a weak metric for rare-event detection?

  \item \MMarks \textbf{[Subjective]} State the objective of k-means and explain why scaling matters.

  \item \MMarks \textbf{[Subjective]} Define state, action, reward, and policy in reinforcement learning.

  \item \MMarks \textbf{[Subjective]} Give one useful semi-supervised learning scenario and one major risk.

  \item \MMarks \textbf{[Subjective]} Give one example of concept drift in banking, education, or security,
    and mention one monitoring method.
\end{enumerate}

\section*{Hard (20 Questions)}
\begin{enumerate}[label=\textbf{H\arabic*.}, leftmargin=*]
  \item \HMarks \textbf{[Subjective]} Design a knowledge-based agent for a smart home security scenario.
    Provide propositional symbols, rules/facts, and one entailment query.

  \item \HMarks \textbf{[Subjective]} Use resolution refutation to show:
    \[
      \{A \rightarrow B,\ B \rightarrow C,\ C \rightarrow D,\ A\} \models D.
    \]

  \item \HMarks \textbf{[Subjective]} Convert to clause form and show Skolemization:
    \[
      \forall x\ (Student(x) \rightarrow \exists y\ Mentors(y,x)).
    \]

  \item \HMarks \textbf{[Subjective]} From
    \[
      \forall x\ (Human(x) \rightarrow Mortal(x)),\quad Human(Socrates)
    \]
    prove $Mortal(Socrates)$ using resolution.

  \item \HMarks \textbf{[Subjective]} Write two FOL translations for ``Every student likes a course''
    and explain the meaning difference.

  \item \HMarks \textbf{[Subjective]} Provide a small KB and a countermodel showing $KB \not\models \alpha$.

  \item \HMarks \textbf{[Subjective]} Design an ML workflow for predicting hostel electricity demand.
    Include features, splits, preprocessing, metrics, and one deployment risk.

  \item \HMarks \textbf{[Subjective]} Give a clear example of data leakage in a student-performance prediction system
    and explain how to fix the workflow.

  \item \HMarks \textbf{[Subjective]} Design an evaluation plan for a healthcare classifier where false negatives are costly.

  \item \HMarks \textbf{[Subjective]} Propose a regression workflow for demand forecasting in energy/utilities.

  \item \HMarks \textbf{[Subjective]} Design a clustering workflow for customer segmentation and explain how you would choose $k$.

  \item \HMarks \textbf{[Subjective]} Design an RL setup for adaptive traffic signal control.

  \item \HMarks \textbf{[Subjective]} Propose a semi-supervised workflow for medical images with few labels and many unlabeled samples.

  \item \HMarks \textbf{[Subjective]} Write a short responsible AI policy note for an AI-based scholarship recommender.

  \item \HMarks \textbf{[Subjective]} Compare one AI/ML application each in banking, security, and healthcare:
    task, metric, risk, and monitoring.

  \item \HMarks \textbf{[Subjective]} Explain why self-driving is difficult using environment properties,
    sensors, and uncertainty.

  \item \HMarks \textbf{[Subjective]} A university chatbot gave incorrect scholarship advice.
    Write a short failure postmortem with causes, impact, and prevention plan.

  \item \HMarks \textbf{[Subjective]} Explain one fairness risk and one privacy risk in an education recommender system,
    plus one mitigation for each.

  \item \HMarks \textbf{[Subjective]} Create a deployment checklist for a student-facing AI application.
    Include privacy, monitoring, fallback, human override, and incident handling.

  \item \HMarks \textbf{[Subjective]} In 12--16 lines, explain how logical reasoning, ML modeling, evaluation metrics,
    and responsible deployment fit together in a complete AI system.
\end{enumerate}

\vfill
\noindent\textit{End of Assessment}

\end{document}
