\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{listings}

\geometry{margin=1in}
\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  urlcolor=blue
}

\setlist{itemsep=0.35em, topsep=0.35em}

\pagestyle{fancy}
\fancyhf{}
\lhead{Elements of AIML (B.Tech)}
\rhead{Assignment: Units 01--05}
\cfoot{\thepage}
\setlength{\headheight}{14pt}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  stringstyle=\color{teal},
  showstringspaces=false,
  columns=fullflexible,
  frame=single
}

\newcommand{\SubmissionDeadline}{April 28, 2026}

\begin{document}

\begin{center}
  {\LARGE \textbf{Assignment -- Units 01--05}}\\[0.25em]
  {\large Elements of AIML}\\[0.25em]
  \normalsize (Introduction to AI, Logic for AI, Machine Learning Basics, ML Types, and AI/ML Applications)
\end{center}

\noindent
\textbf{Instructor:} Tofik Ali \hfill \textbf{Submission Deadline:} \SubmissionDeadline\\
\textbf{Student Name:} \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_ \hfill \textbf{SAP-ID:} \_\_\_\_\_\_\_\_\_\_\_\_\\

\vspace{0.6em}
\hrule
\vspace{0.8em}

\textbf{Instructions}
\begin{itemize}[leftmargin=*]
  \item Answer \textbf{all} questions. This assignment is cumulative and covers \textbf{Unit 01 to Unit 05}.
  \item Submission format: submit one ZIP containing:
    \begin{itemize}[leftmargin=*]
      \item \texttt{notebook.ipynb} (typed answers with equations/diagrams where useful),
      \item \texttt{README.md} (1-page summary of what you revised and references used),
      \item optional \texttt{assets/} folder for tables, screenshots, or diagrams.
    \end{itemize}
  \item Use your own examples whenever the question is open-ended.
  \item For logic questions, show major steps clearly (truth tables, CNF conversion, resolution, substitutions).
  \item For ML questions, explain data assumptions, metrics, and why your workflow avoids leakage.
  \item Write clearly enough that another student could study from your solution.
\end{itemize}

\vspace{0.6em}

\section*{Easy (20 Questions)}
\begin{enumerate}[label=E\arabic*., leftmargin=*]
  \item \textbf{Define AI:} Give a 2--3 line definition of Artificial Intelligence in your own words.
    Mention one key goal such as perception, planning, language, or decision-making.

  \item \textbf{AI applications:} List five AI applications from different domains.
    For each, write one line in the form \emph{input} $\rightarrow$ \emph{output}.

  \item \textbf{AI vs ML vs Data Science:} Write 6--10 lines comparing AI, Machine Learning, and Data Science.
    Give one example project for each.

  \item \textbf{Technique categories:} Name three broad AI technique families and give one example of each:
    search/planning, logic/reasoning, and learning.

  \item \textbf{Success criteria:} For each system below, write two success criteria beyond accuracy:
    \begin{enumerate}[label=(\alph*), leftmargin=*]
      \item medical screening support,
      \item loan approval decision support.
    \end{enumerate}

  \item \textbf{Agent basics:} Define an \textbf{agent}. Give one example and list its sensors and actuators.

  \item \textbf{PEAS:} Write a PEAS description for a spam filter.

  \item \textbf{Environment properties:} Classify \textbf{chess} and an \textbf{online shopping recommender}
    using the following properties where appropriate: observable, deterministic, episodic/sequential, static/dynamic.
    Justify briefly.

  \item \textbf{Rationality:} Define a rational agent in 3--5 lines.
    Explain the difference between a performance measure and a reward.

  \item \textbf{Limitations:} List four deployment-time limitations of AI systems.
    For each, write one practical mitigation.

  \item \textbf{PL translation:} Let $P$ = ``Smoke is detected'' and $Q$ = ``Raise an alert''.
    Translate: ``If smoke is detected then raise an alert, and smoke is detected.''
    Write the formula and state what it entails.

  \item \textbf{Truth table:} Write the truth table for $P \rightarrow Q$ and state when it is false.

  \item \textbf{Validity and satisfiability:} Define \textbf{valid} and \textbf{satisfiable}.
    Give one example sentence for each.

  \item \textbf{Entailment vs implication:} Explain in 6--10 lines the difference between $KB \models \alpha$ and $KB \rightarrow \alpha$.

  \item \textbf{FOL basics:} In $Studies(Student, Course)$ identify the predicate symbol, constants, and variables (if any).
    Then write one quantified sentence using the same predicate.

  \item \textbf{Dataset vocabulary:} Define \textbf{dataset}, \textbf{instance}, \textbf{feature}, and \textbf{label}
    with one small example.

  \item \textbf{Data splitting:} Explain the roles of train, validation, and test sets in 5--7 lines.

  \item \textbf{Leakage:} Give one concrete example of data leakage and explain why it causes an unreliable evaluation.

  \item \textbf{PCA intuition:} State what PCA optimizes and explain in 3--5 lines why standardization is often required before PCA.

  \item \textbf{Learning task types:} Distinguish regression, classification, and clustering.
    Give one example problem for each.
\end{enumerate}

\section*{Medium (20 Questions)}
\begin{enumerate}[label=M\arabic*., leftmargin=*]
  \item \textbf{PEAS design:} Create a PEAS description for an online exam proctoring system.
    Include at least five items for performance measure and sensors.

  \item \textbf{Environment analysis:} Consider a smart thermostat in a hostel room.
    Classify its environment using at least five properties and justify each.

  \item \textbf{Agent comparison:} Compare simple reflex, model-based reflex, goal-based, utility-based, and learning agents in a table.
    Add one example task for each.

  \item \textbf{Choosing an agent:} You must design an elevator controller for a multi-floor academic block.
    Choose one agent type and justify your choice in 8--12 lines.

  \item \textbf{Model checking:} Let $KB=\{P \rightarrow Q,\ Q \rightarrow R,\ P\}$.
    Check whether $KB \models R$ using a truth table or model enumeration.

  \item \textbf{CNF conversion:} Convert $\lnot(P \rightarrow Q)$ to CNF and show the main steps.

  \item \textbf{CNF with biconditional:} Convert
    \[
      \lnot(P \leftrightarrow Q) \rightarrow (R \lor P)
    \]
    to CNF. Show the major transformations.

  \item \textbf{Resolution proof:} Using resolution refutation, prove
    \[
      \{P \rightarrow Q,\ Q \rightarrow R,\ P\} \models R.
    \]

  \item \textbf{English to FOL:} Domain: students and courses. Translate into FOL:
    \begin{enumerate}[label=(\alph*), leftmargin=*]
      \item Every student studies at least one course.
      \item Some course is studied by every student.
    \end{enumerate}

  \item \textbf{Unification:} Find an MGU if it exists:
    \[
      Knows(John,x) \quad \text{and} \quad Knows(y,Bill).
    \]

  \item \textbf{Occurs-check:} Explain why unifying $x$ with $f(x)$ should fail.
    State one issue it can cause in inference.

  \item \textbf{Dataset design:} You want to predict hostel electricity usage for the next day.
    Propose at least six features, one target variable, and one preprocessing step.

  \item \textbf{Split design:} For a dataset of 300 student records with an imbalanced final outcome label,
    propose a train/validation/test split and explain why stratification matters.

  \item \textbf{Cross-validation:} Explain when k-fold cross-validation is better than one fixed split.
    Mention one limitation of cross-validation.

  \item \textbf{Explained variance:} Suppose PCA gives explained variance ratios
    \[
      [0.52, 0.21, 0.12, 0.08, 0.07].
    \]
    How many components are required to retain at least 80\% variance and at least 90\% variance?

  \item \textbf{PCA vs LDA vs ICA:} Compare these three methods conceptually.
    Mention the main objective of each and one suitable use-case.

  \item \textbf{Confusion matrix metrics:} A classifier gives:
    TP = 45, FP = 15, FN = 5, TN = 35.
    Compute accuracy, precision, recall, and F1-score.

  \item \textbf{MAE vs RMSE:} Explain the difference between MAE and RMSE.
    When is RMSE usually preferred?

  \item \textbf{k-means basics:} State the k-means objective function in words and write two practical pitfalls of using k-means.

  \item \textbf{Learning paradigms:} Contrast supervised, unsupervised, semi-supervised,
    and reinforcement learning. Give one example for each.
\end{enumerate}

\section*{Hard (20 Questions)}
\begin{enumerate}[label=H\arabic*., leftmargin=*]
  \item \textbf{Full case design:} Design an AI system for waste collection optimization in a city.
    Provide the agent description, PEAS, success criteria, and three major risks with mitigations.

  \item \textbf{Self-driving analysis:} Explain why self-driving is a difficult AI task.
    Classify the environment using at least six properties and justify each.

  \item \textbf{Learning agent decomposition:} For a news recommendation system, describe the performance element,
    learning element, critic, and problem generator. Draw a simple block diagram if useful.

  \item \textbf{Knowledge-based agent:} Design a propositional knowledge base for a smart home security scenario.
    Provide at least six symbols, six rules/facts, and one query answered through entailment.

  \item \textbf{Integrated reasoning:} A campus assistant must decide whether to send an emergency alert.
    Build a propositional KB with at least six sentences involving alarm, smoke, system fault, and manual override.
    Ask two queries and answer them using model checking or resolution.

  \item \textbf{Resolution chain:} Prove using resolution refutation:
    \[
      \{A \rightarrow B,\ B \rightarrow C,\ C \rightarrow D,\ A\} \models D.
    \]

  \item \textbf{Clause form:} Convert to clause form and show Skolemization:
    \[
      \forall x\ (Student(x) \rightarrow \exists y\ Mentors(y,x)).
    \]

  \item \textbf{FOL resolution proof:} From
    \[
      \forall x\ (Human(x) \rightarrow Mortal(x)), \quad Human(Socrates)
    \]
    prove $Mortal(Socrates)$ using resolution.

  \item \textbf{Quantifier scope ambiguity:} Write two FOL translations for
    ``Every student likes a course.''
    Explain the meaning difference clearly.

  \item \textbf{Countermodel:} Give a small KB and show a countermodel that proves $KB \not\models \alpha$.
    Write the interpretation or assignment explicitly.

  \item \textbf{Full ML pipeline:} Design a machine learning workflow for predicting student dropout risk.
    Include data collection, splitting, preprocessing, validation, metric choice, and post-deployment monitoring.

  \item \textbf{PCA workflow:} You have a 30-feature sensor dataset for equipment monitoring.
    Explain how you would apply PCA end-to-end, choose $k$, and decide whether PCA is helping.

  \item \textbf{Regression case study:} Propose a regression pipeline for energy demand forecasting.
    Include feature ideas, metric choice, train/validation/test strategy, and one deployment risk.

  \item \textbf{Classification case study:} Design a healthcare screening classifier where false negatives are costly.
    Explain metric choice, threshold setting, and one monitoring plan after deployment.

  \item \textbf{Clustering case study:} A retailer wants customer segments.
    Propose a clustering workflow including preprocessing, choice of $k$, validation, and business interpretation.

  \item \textbf{Reinforcement learning design:} Design an RL setup for adaptive traffic signal control.
    Define state, action, reward, policy objective, and one exploration challenge.

  \item \textbf{Semi-supervised design:} A medical imaging project has few labeled scans but many unlabeled ones.
    Propose a semi-supervised workflow and explain how you would control pseudo-labeling errors.

  \item \textbf{Responsible AI policy note:} Write a one-page policy note for an AI-based scholarship recommender.
    Include fairness, privacy, transparency, accountability, and human override.

  \item \textbf{Cross-domain comparison:} Compare AI/ML use in banking, security, and healthcare.
    For each, identify one task, one important metric, one major risk, and one monitoring need.

  \item \textbf{Course reflection:} In 14--20 lines, explain how logical reasoning, ML evaluation, and responsible deployment
    fit together in real AI systems.
\end{enumerate}

\vfill
\noindent\textit{End of Assignment}

\end{document}
