\documentclass[11pt]{article}

\input{../../../_shared/notes-common.tex}

\title{Elements of AIML Lab\\Experiment 10: Visualization (Student Handout — Before Submission)}
\author{Tofik Ali}
\date{\today}

\begin{document}
\maketitle
\tableofcontents

\section{About this handout}
This is the \textbf{pre-submission} student handout.
It intentionally provides only \textbf{requirements} and \textbf{hints}.
After submission, you may receive a separate \textbf{reference} document for cross-checking.

\section{Objective}
\begin{itemize}[leftmargin=*]
  \item Create a small portfolio of plots and write observations for each.
  \item Demonstrate good visualization hygiene (titles, labels, readable scales).
\end{itemize}

\section{What you must do (minimum requirements)}
\begin{enumerate}[leftmargin=*]
  \item Choose a dataset (from previous labs preferred) and record its source/link.
  \item Create and save at least \textbf{five} plots, including:
  \begin{itemize}[leftmargin=*]
    \item one numeric distribution plot (histogram/KDE),
    \item one boxplot,
    \item one categorical count plot (bar/count plot),
    \item one relationship plot (scatter),
    \item one correlation-style plot (heatmap) for numeric features.
  \end{itemize}
  \item Every plot must have a title and labeled axes (and legend if needed).
  \item For each plot, write 2--3 lines of observation (what it shows and what it suggests).
\end{enumerate}

\section{Hints (read only if stuck)}
\begin{itemize}[leftmargin=*]
  \item Use Matplotlib + Seaborn; save plots as \texttt{.png} at readable resolution.
  \item Choose plot types based on variable types (numeric vs categorical vs time).
  \item If you have too many categories, plot top-$k$ categories and group the rest as ``Other''.
\end{itemize}

\section{What to submit (MANDATORY)}
Submit a single zipped folder containing:
\begin{itemize}[leftmargin=*]
  \item \texttt{notebook.ipynb}: code + plots + observations.
  \item \texttt{README.md}: dataset link/source and a list of plots with key takeaways.
  \item \texttt{plots/} (folder): exported \texttt{.png} files.
\end{itemize}
Dataset rule: do not upload large datasets; provide a link/source instead.

\section{Submission checklist}
\begin{itemize}[leftmargin=*]
  \item Five required plot types included and saved.
  \item Titles and axis labels present on all plots.
  \item Observations written for each plot.
  \item Notebook + README included and readable.
\end{itemize}

\section{Common mistakes}
\begin{itemize}[leftmargin=*]
  \item Missing labels/titles (plots become unusable).
  \item Wrong plot choice for variable type.
  \item Saving low-resolution plots that cannot be read.
\end{itemize}

\notesources{\AIMLLabSources}
\end{document}
