\documentclass[11pt]{article}

\input{../../../_shared/notes-common.tex}

\title{Elements of AIML Lab\\Experiment 07: Scaling \& One-Hot Encoding (Student Handout — Before Submission)}
\author{Tofik Ali}
\date{\today}

\begin{document}
\maketitle
\tableofcontents

\section{About this handout}
This is the \textbf{pre-submission} student handout.
It intentionally provides only \textbf{requirements} and \textbf{hints}.
After submission, you may receive a separate \textbf{reference} document for cross-checking.

\section{Objective}
\begin{itemize}[leftmargin=*]
  \item Apply feature scaling for numeric features and one-hot encoding for categorical features.
  \item Demonstrate a leakage-safe preprocessing approach (train-only fit).
\end{itemize}

\section{What you must do (minimum requirements)}
\begin{enumerate}[leftmargin=*]
  \item Use the dataset from Experiment~04/05 (preferred) or another approved dataset.
  \item Identify numeric vs categorical columns.
  \item Apply \textbf{one} scaling method to numeric columns (standardization or min-max normalization).
  \item Apply \textbf{one-hot encoding} to categorical columns.
  \item Demonstrate leakage-safe practice: preprocessing must be fit on \textbf{training} data and applied to test/validation (use a pipeline approach).
  \item Report the final feature count after encoding and write 3--5 observations.
\end{enumerate}

\section{Hints (read only if stuck)}
\begin{itemize}[leftmargin=*]
  \item Common tools: \texttt{StandardScaler}, \texttt{MinMaxScaler}, \texttt{OneHotEncoder}.
  \item Pipeline helpers: \texttt{ColumnTransformer} and \texttt{Pipeline}.
  \item For unseen categories in test data, \texttt{handle\_unknown="ignore"} avoids errors.
  \item Do not encode ID-like columns (they create useless features and can leak identity).
\end{itemize}

\section{What to submit (MANDATORY)}
Submit a single zipped folder containing:
\begin{itemize}[leftmargin=*]
  \item \texttt{notebook.ipynb}: preprocessing steps and before/after shapes (feature counts).
  \item \texttt{README.md}: dataset link/source, scaling method, encoding method, and leakage-safe approach summary.
  \item \texttt{assets/} (optional): any saved outputs/plots.
\end{itemize}
Dataset rule: do not upload large datasets; provide a link/source instead.

\section{Submission checklist}
\begin{itemize}[leftmargin=*]
  \item Numeric vs categorical columns identified.
  \item Scaling + one-hot encoding applied and documented.
  \item Leakage-safe pipeline (train-only fit) demonstrated.
  \item Notebook + README included and readable.
\end{itemize}

\section{Common mistakes}
\begin{itemize}[leftmargin=*]
  \item Scaling/encoding before the train/test split (leakage).
  \item One-hot encoding an ID column.
  \item Not reporting how many features were created after encoding.
\end{itemize}

\notesources{\AIMLLabSources}
\end{document}
